---
title: 'Distributed lag models (DLMs) for retrospective cohort data: sample code'
author: "Jennifer F. Bobb"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: true
    theme: cerulean
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this file, we provide sample code to fit the subcohort and full-cohort distributed lag models (DLMs) considered in the paper 'Distributed lag models for retrospective cohort data with application to a study of built environment and body weight'. We run the models on a single generated dataset for one of the simulation scenarios considered in the paper. We focus on the main scenario considered of a linear exposure-response function at each lag. Each of the fitted models assumes a linear exposure-response function across all lags, except we also include a one model that allows for a nonlinear exposure-response function (referred to as a distributed lag nonlinear model [DLNM]) to illustrate code for this method.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(magrittr)
library(tidyr)

library(splines)
library(dlnm)

library(mice)
library(miceafter)
library(mitools)

library(ggplot2)

source("functions/generate_cohort_data.R")
source("functions/fit_models.R")
```

## Models to compare

We consider the same set of models as from the simulation study, with the following modifications:

- for the multiple imputation with chained equations (MICE) procedure, we impute 10 rather than 25 datasets and use a maximum of 25 rather than 50 iterations (cycles) for each dataset, to make this sample code easier to run
- we exclude subcohort-3 to reduce the number of plots shown
- we also include a distributed lag nonlinear model (DLNM) based on the full-cohort model with multiple imputation, to provide code for this method. This method is implemented to model the exposure-response function at each lag using natural cubic splines with 3 degrees of freedom (DF, referred to as `df_splX` in the code)

```{r}
impute_params <- tibble::tibble(maxit = 25, m = 10)
model_info <- bind_rows(
    #tibble(method = "subcohort", prior_years = 3, df_spl = NA),
    tibble(method = "subcohort", prior_years = 6, df_spl = c(3, NA)),
    tibble(method = "subcohort", prior_years = 10, df_spl = c(3, 6, NA)),
    tibble(method = "subcohort", prior_years = 12, df_spl = c(3, 6, 9, NA)),
    tibble(method = "full", df_spl = c(3, 6, 9, NA)),
    tibble(method = "full-imp", df_spl = c(3, 6, 9, NA)),
    tibble(method = "full-imp-DLNM", df_spl = c(3, 6, 9, NA), df_splX = 3)
)
```

## Generate dataset from a particular simulation scenario

Specify the parameters for the simulation. Here we generate data from a linear exposure-response function at each lag.

```{r}
## scenario for autocorrelation of lagged exposures
expos_rho_sel <- 0.75
#expos_rho_sel <- 0.97

## scenario of true distributed lag coefficient function
#DL_scen_sel <- "constant"
DL_scen_sel <- "proximal"
#DL_scen_sel <- "distal"

## Proportion in cohort having different lengths of exposure history
prop_subcohorts <- structure(list(n_prior_expos = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 
11, 12), prop = c(0.17, 0.17, 0.13, 0.11, 0.09, 0.07, 0.06, 0.05, 
0.04, 0.04, 0.04, 0.03)), row.names = c(NA, -12L), class = c("tbl_df", 
"tbl", "data.frame"))

#n_obs <- 2000
n_obs <- 1000 ## sample size
n_cohorts <- 12 ## maximum number of prior exposures

## Distributed lag coefficient function scenarios
cfs <- c(2, 1, 0.5, 0.25, rep(0, n_cohorts-4)) 
DL_fun_scen <- bind_rows(
    tibble(DL_scen = "constant", DL_coef = rep(0.3, length(cfs)) %>% list()),
    tibble(DL_scen = "proximal", DL_coef = cfs %>% list()),
    tibble(DL_scen = "distal",  DL_coef = rev(cfs) %>% list())
) %>%
    filter(DL_scen == DL_scen_sel)

## all parameter values for current data generation
params <- tibble(
    n_obs = n_obs,
    n_cohorts = n_cohorts,
    expos_rho = expos_rho_sel, ## correlation of lagged exposures
    sigsq_y_true = c(11)^2, ## residual variance in outcome generating model
    intercept = 0, ## intercept in outcome generating model
    expos_gen_info = tibble(expos_mean = 10, expos_SD = 8) %>% list() ## mean and SD of lagged exposures,
) %>%
    bind_cols(DL_fun_scen)
```

For illustration, here we generate a single dataset from one set of parameter values. To save computation time for this illustration we generate a dataset of size `r n_obs` (in the paper we used 2000). The parameter values we use are:

- We set the autocorrelation parameter for lagged exposures (rho) to be `r expos_rho_sel`
- We set the true distributed lag coefficient to be the '`r DL_scen_sel`' function


```{r}
## Information on exposure variable names in datasets
expos_names_prefix <- "expos" ## exposure at lag 1 will be named 'expos1', etc.
expos_var_names <- paste0(expos_names_prefix, 1:n_cohorts)

## First, generate components used in data generating models
seed_dataset <- 123
set.seed(seed_dataset)
e <- params$expos_gen_info[[1]]
df_comps <- params %$% gen_data_comps(
    n_obs = n_obs, n_cohorts = n_cohorts, 
    expos_mean = e$expos_mean, 
    expos_SD = e$expos_SD, 
    expos_rho = expos_rho_sel, 
    prop_cohorts = prop_subcohorts$prop, 
    sigsq_y_true = sigsq_y_true
)

## Now generate outcome data and missingness from components
dset <- params %$% gen_data_from_comps(
    df_comps = df_comps, DL_coef = DL_coef[[1]]
) %>%
    select(-eps, -mu)
```

## Impute complete exposure data for full-cohort using multiple imputation (MI)

```{r}
seed_dataset2 <- seed_dataset + 1

## Uses the mice R package
to_impute <- dset %>%
    select(y, all_of(expos_var_names))

time_imp0 <- Sys.time()
imp <- mice(
    to_impute, 
    maxit = impute_params$maxit, 
    m = impute_params$m, 
    printFlag = FALSE, 
    remove.collinear = FALSE ## some iterations were leading to errors in the subsequent DLM as expos12 was not being imputed due to colinearity
)
time_imp1 <- Sys.time()
#difftime(time_imp1, time_imp0) %>% capture.output() %>% message()
## took ~12.5 secs for n_obs=1000
imputed_data <- imp
```


## Fit models to the simulated dataset

The following code loops through all the methods to apply to the simulated dataset.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
res <- NULL
full_time0 <- Sys.time()
for(i in 1:nrow(model_info)) {
    res0 <- model_info[i, ]
    seed_iter <- seed_dataset
    df_spl <- res0$df_spl
    df_splX <- res0$df_splX
    
    if (res0$method == "subcohort") {
        form0 <- paste0("y ~ 1") ## covariate adjustment (here no covariates)
        prior_years_sel <- res0$prior_years ## years of prior exposures available for the subcohort
        expos_var_names_subcohort <- expos_var_names[1:prior_years_sel]
        dat_subcohort <- dset %>%
            filter(n_prior_expos >= prior_years_sel) ## identify subcohort
        
        if (!is.na(df_spl)) { ## constrained DLM
            fit <- fit_DLM_constrained(df_spl = df_spl, max_lag = prior_years_sel, data = dat_subcohort, expos_var_names = expos_var_names_subcohort, form0 = form0)
        } else { ## unconstrained DLM
            fit <- fit_unconstr_dlm(expos_var_names = expos_var_names_subcohort, data = dat_subcohort, form0 = form0)
        }
        
    } else if (res0$method == "full") {
        ## create interaction terms needed to fit full-cohort model using observed cohort data
        dset_int_terms <- set_up_lagged_dset(dset, expos_names_prefix = expos_names_prefix)
        
        ## exposure variable names from interaction model
        expos_var_names_anal <- attr(dset_int_terms, "expos_var_names_anal")
        
        if (!is.na(df_spl)) { ## constrained DLM
            fit <- fit_DLM_constrained(df_spl = df_spl, max_lag = n_cohorts, data = dset_int_terms, expos_var_names = expos_var_names_anal, form0 = "y ~ cohort")
        } else { ## unconstrained DLM
            fit <- fit_unconstr_dlm(expos_var_names = expos_var_names_anal, data = dset_int_terms, form0 = "y ~ cohort")
        }
    } else if (res0$method == "full-imp") {
        
        if (!is.na(df_spl)) { ## constrained DLM
            fit <- fit_DLM_constrained_MI(df_spl = df_spl, max_lag = n_cohorts, imputed_data = imputed_data, expos_var_names = expos_var_names, form0 = form0)
        } else { ## unconstrained DLM
            fit <- try(fit_unconstr_dlm_MI(imputed_data = imputed_data, expos_var_names = expos_var_names, form0 = form0))
        }
    } else if (res0$method == "full-imp-DLNM") {
        
        fit <- fit_DLM_constrained_MI_splX(
            df_spl = ifelse(is.na(df_spl), n_cohorts, df_spl), ## setting the number of DF to be the number of exposures is equivalent to the constrained DLM, 
            max_lag = n_cohorts, imputed_data = imputed_data, expos_var_names = expos_var_names, form0 = form0, df_splX = df_splX)
    }
    
    ## Add 95% Wald CI
    fit_addci <- fit %>% 
        mutate(lb = Estimate - 1.96*SE,
               ub = Estimate + 1.96*SE)
    
    ## Store lag-specific and cumulative estimates
    res_iter <- res0 %>% 
        mutate(seed_iter = seed_iter,
               lag_ests = list(fit_addci %>% filter(!is.na(lag))),
               cum_est = list(fit_addci %>% filter(is.na(lag))))
    res %<>% bind_rows(res_iter)

}
full_time1 <- Sys.time()
#difftime(full_time1, full_time0) ## 1.2 secs to run all the methods

## Add method names to use for plotting
res_all <- res %>%
    mutate(
        method_base = ifelse(!is.na(prior_years), paste0(method, prior_years), method),
        method_base_fac = factor(method_base, levels = unique(method_base))
    ) %>%
    mutate(
        df_spl = as.factor(ifelse(is.na(df_spl), "NA", df_spl)) ## edit levels for plotting purposes
    )
```

## Parameter estimates across models

In the figures below, the subcohort models include individuals with at least 6, 10, or all 12 years of exposure history. Full-cohort models use observed data ("full") or imputed data ("full-imp"). Based on the results presented in the paper, the "full" method is not recommended for use due to its poor performance (e.g., bias in estimating the true DLM coefficient function). The parameter `df_spl` corresponds to the number of degrees of freedom (DF) of the lag function; a value of "NA" corresponds to the unconstrained DLM.

### Lag-specific estimates
```{r, fig.width=8, fig.height=5}
lag_ests_true <- tibble(lag = 1:n_cohorts, beta_true = params$DL_coef[[1]])

lag_ests <- res_all %>% 
    unnest(lag_ests) %>% 
    left_join(lag_ests_true, by = "lag")

lag_ests %>% 
    ggplot(aes(lag, Estimate, ymin = lb, ymax = ub)) +
    geom_line(aes(lag, beta_true), col = "black", lwd = 0.7) +
    geom_pointrange(aes(shape = df_spl, col = df_spl), position = position_dodge(width = 0.5), cex = 0.3) +
    facet_wrap(~ method_base_fac, labeller = labeller(.cols = label_parsed, .multi_line = FALSE)) +
    theme_bw() +
    xlab("Lag") + ylab("Estimate") +
    scale_x_continuous(breaks = (c(2,4,6,8,10,12)))
```

### Cumulative estimates
```{r, fig.width=7, fig.height=2.75}
cum_ests_true <- sum(params$DL_coef[[1]])

cum_ests <- res_all %>% 
    unnest(cum_est)

cum_ests %>% 
    ggplot(aes(method_base_fac, Estimate, ymin = lb, ymax = ub)) +
    geom_pointrange(aes(shape = df_spl, col = df_spl), position = position_dodge(width = 0.5), cex = 0.3) +
    geom_hline(yintercept = cum_ests_true, linetype = 2, col = "black") +
    theme_bw() +
    xlab("Method") + ylab("Estimate")

```



